---
title: "Irish Public Hospital Waiting List Analysis"
subtitle: "NTPF Open Data 2024–2026 | Health Data Science Portfolio"
author: "Conor McCool"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: kable
    fig_width: 12
    fig_height: 7
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 12,
  fig.height = 7,
  fig.align  = "center",
  dpi        = 150
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)
library(forecast)
library(cluster)
library(factoextra)
library(broom)
library(zoo)
library(kableExtra)
```

---

# Introduction {.tabset}

## Dataset Overview

This report analyses the **National Treatment Purchase Fund (NTPF) Outpatient Waiting List Open Data**, covering all public hospital outpatient waiting lists in Ireland from January 2024 to the most recently published month of 2026. The NTPF is an independent public body that collects, validates and publishes monthly waiting list data for every consultant-led outpatient service across Ireland's public hospital system.

The dataset is stratified by:

- **Specialty** — over 50 clinical specialties including Cardiology, Neurology, Respiratory Medicine, and Gastroenterology
- **Adult / Child** — separate reporting for adult and paediatric patients
- **Waiting time band** — patients are bucketed into four bands: 0–6 months, 6–12 months, 12–18 months, and 18+ months
- **Archive Date** — a monthly snapshot, published on the last Friday of each month

This data is directly analogous to the kind of population health surveillance work carried out by the **HSE National Health Intelligence Unit (NHIU)** — particularly in relation to the Chronic Disease Management (CDM) dataset, where understanding waiting volumes, time distributions and specialty-level trends is central to health service planning.

## Goals of This Analysis

The analysis pursues five overarching objectives:

1. **Data Quality** — Establish the reliability, completeness and internal consistency of the dataset before any analytical conclusions are drawn. In a health policy context, decisions made on flawed data carry real consequences.

2. **Descriptive Intelligence** — Characterise the current state of outpatient waiting lists: which specialties carry the highest burden, how the time band composition has shifted, and what disparities exist between adult and paediatric patients.

3. **Temporal Modelling** — Decompose and forecast the national waiting list trend using time series methods (STL decomposition and ARIMA modelling), enabling evidence-based projections of future demand.

4. **Segmentation** — Cluster specialties by their waiting list behaviour using k-means, identifying groups that share structural characteristics and may therefore respond to similar policy interventions.

5. **Statistical Validation** — Test whether year-on-year changes are statistically significant and quantify the drivers of waiting list growth through regression analysis.

## Data Challenges

Working with administrative health data of this kind presents several well-documented challenges that any rigorous analysis must address:

**Structural challenges:**

- **Mixed date formats** across years — the `ArchiveDate` column uses `dd/mm/yyyy` formatting that varies across years and must be parsed carefully before any date arithmetic.
- **Column name inconsistency** — raw column names contain spaces, slashes and special characters that must be standardised before joining or pivoting.
- **Duplicate records** — each monthly file may contain both adult and child rows for the same specialty, requiring deduplication logic before pivot operations.

**Analytical challenges:**

- **Non-normality** — waiting list counts are right-skewed with occasional large outliers, meaning standard parametric tests (t-tests) are inappropriate. Non-parametric alternatives (Wilcoxon signed-rank) are used throughout.
- **Small specialty volumes** — the NTPF suppresses counts below 20 patients (aggregated under "Small Volume") to preserve confidentiality via Statistical Disclosure Control. These rows require careful handling.
- **Short time series** — with only 28 months of data across 2024–2026, ARIMA forecasts carry wider confidence intervals than a longer series would produce. Results should be interpreted with this in mind.
- **No activity data** — the NTPF explicitly notes it does not collect data on the number of patients *treated or removed* each month. Waiting list totals therefore reflect accumulation, not throughput.

---

# Section 1: ETL — Data Loading {.tabset}

## Approach

Data is loaded directly from the NTPF's publicly accessible CSV endpoints — one file per calendar year, each containing monthly snapshots for all specialties. This mirrors the kind of scheduled pipeline ingestion the NHIU operates against datasets like the Chronic Disease Management register.

Using `imap_dfr()` from `purrr`, all three years are loaded in a single vectorised call. The named list structure means adding a new year requires only one additional line — the function handles the rest automatically.

```{r etl-load}
load_ntpf_op <- function(year, url) {
  read_csv(url, show_col_types = FALSE) %>%
    mutate(Year = as.integer(year))
}

urls <- list(
  `2024` = "https://www.ntpf.ie/app/uploads/2025/01/OpenData_OPNational02_2024-1.csv",
  `2025` = "https://www.ntpf.ie/app/uploads/2026/01/OpenData_OPNational02_2025.csv",
  `2026` = "https://www.ntpf.ie/app/uploads/2026/05/OpenData_OPNational02_2026.csv"
)

# imap passes (.x = value, .y = name), so lambda maps correctly
op_raw <- imap_dfr(urls, ~ load_ntpf_op(year = .y, url = .x))
```

## Raw Data Shape

```{r etl-shape}
cat("Rows loaded:", nrow(op_raw), "\n")
cat("Columns:", ncol(op_raw), "\n")
cat("Years present:", paste(unique(op_raw$Year), collapse = ", "), "\n")
glimpse(op_raw)
```

---

# Section 2: Data Quality Assessment {.tabset}

## Overview

Before any analytical step, a thorough quality assessment is essential. In a health intelligence setting, acting on flawed data has real downstream consequences — incorrect waiting list estimates could misallocate resources or misrepresent performance to the Department of Health. Three checks are performed: **completeness**, **internal consistency** and **temporal integrity**.

## Column Name Cleaning

```{r dqa-rename}
op_raw <- op_raw %>%
  rename_with(
    ~ str_replace_all(., "[^a-zA-Z0-9]", "_") %>%
      str_replace_all("_+", "_") %>%
      str_remove("_$") %>%
      str_remove("^_")
  )

cat("Cleaned column names:\n")
print(names(op_raw))
```

## Completeness Report

```{r dqa-completeness}
completeness <- op_raw %>%
  summarise(across(everything(), ~ round(mean(is.na(.)) * 100, 2))) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "Pct_Missing") %>%
  arrange(desc(Pct_Missing))

completeness %>%
  kable(caption = "Missingness (%) by Column", col.names = c("Column", "% Missing")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  row_spec(which(completeness$Pct_Missing > 5), bold = TRUE, color = "white", background = "#E8491D")
```

```{r dqa-flag}
high_missing <- completeness %>% filter(Pct_Missing > 5)
if (nrow(high_missing) > 0) {
  cat("⚠️  High missingness detected in:", paste(high_missing$Column, collapse = ", "))
} else {
  cat("✅  All columns within acceptable missingness threshold (<5%)")
}
```

## Date Parsing & Feature Engineering

```{r dqa-parse}
op_clean <- op_raw %>%
  mutate(
    Archive_Date = as.Date(parse_date_time(ArchiveDate, orders = c("dmy", "ymd"))),
    Month        = floor_date(Archive_Date, "month"),
    Quarter      = quarter(Archive_Date, with_year = TRUE),
    Month_Num    = month(Archive_Date),
    Year_Factor  = as.factor(Year)
  ) %>%
  rename(
    Specialty = Speciality,
    M0_6      = `0_6_Months`,
    M6_12     = `6_12_Months`,
    M12_18    = `12_18_Months`,
    M18_Plus  = `18_Months`
  ) %>%
  filter(!is.na(Total), !is.na(Archive_Date))

cat("Rows after cleaning:", nrow(op_clean), "\n")
cat("Date range:", format(min(op_clean$Archive_Date), "%B %Y"),
    "to", format(max(op_clean$Archive_Date), "%B %Y"), "\n")
cat("Specialties:", n_distinct(op_clean$Specialty), "\n")
```

## Consistency & Integrity Checks

```{r dqa-consistency}
# Check: do the four time bands sum to the Total?
op_clean <- op_clean %>%
  mutate(
    Band_Sum     = rowSums(across(c(M0_6, M6_12, M12_18, M18_Plus)), na.rm = TRUE),
    Total_Check  = abs(Band_Sum - Total),
    Inconsistent = Total_Check > 5
  )

n_inc <- sum(op_clean$Inconsistent, na.rm = TRUE)
pct_inc <- round(n_inc / nrow(op_clean) * 100, 2)
cat("Inconsistent band-sum rows:", n_inc, "(", pct_inc, "% of data)\n")

# Check: duplicate month/specialty/group combinations
duplicates <- op_clean %>%
  count(Archive_Date, Specialty, Adult_Child) %>%
  filter(n > 1)

if (nrow(duplicates) == 0) {
  cat("✅  Temporal integrity check passed — no duplicate records\n")
} else {
  cat("⚠️  Duplicate records found:", nrow(duplicates), "\n")
  print(duplicates)
}

op_clean <- op_clean %>% select(-Band_Sum, -Total_Check, -Inconsistent)
```

---

# Section 3: Exploratory Data Analysis {.tabset}

## Overview

With a validated, clean dataset we begin exploratory analysis. EDA serves three purposes here: understanding the **distributional shape** of the data (which informs modelling choices), identifying the **highest-burden specialties** at the most recent snapshot, and characterising **how the waiting list composition has shifted over time** across the four waiting bands.

## Distribution by Time Band

Each histogram represents the spread of monthly patient counts across all specialty-month observations for that band. The pronounced right skew in every band — with a small number of very high-volume specialty months pulling the tail — is a key reason parametric tests are inappropriate for this data and why non-parametric methods are used later.

```{r eda-distribution, fig.height=8}
op_clean %>%
  select(M0_6, M6_12, M12_18, M18_Plus) %>%
  pivot_longer(everything(), names_to = "Band", values_to = "Patients") %>%
  mutate(Band = factor(Band,
                       levels = c("M0_6", "M6_12", "M12_18", "M18_Plus"),
                       labels = c("0–6 Months", "6–12 Months",
                                  "12–18 Months", "18+ Months"))) %>%
  filter(!is.na(Patients), Patients > 0) %>%
  ggplot(aes(x = Patients, fill = Band)) +
  geom_histogram(bins = 40, alpha = 0.85, colour = "white") +
  facet_wrap(~ Band, scales = "free", ncol = 2) +
  scale_x_continuous(labels = comma) +
  scale_fill_manual(values = c("#41B6E6", "#005EB8", "#003087", "#E8491D")) +
  labs(
    title    = "Distribution of Patient Counts by Waiting Time Band",
    subtitle = "Each bar = one specialty–month observation | All years combined",
    x        = "Patients Waiting", y = "Frequency"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold", size = 12))
```

## Top 10 Specialties — Latest Snapshot

The bar chart below shows the ten highest-volume outpatient waiting lists at the most recently published date. Colour encodes the **proportion of that specialty's list waiting 18 months or more** — a critical metric for health equity. Specialties appearing in deep orange carry both high volume *and* high proportions of long waiters, representing the most acute pressure points in the system.

```{r eda-top10, fig.height=7}
latest_date <- max(op_clean$Archive_Date)

top10 <- op_clean %>%
  filter(Archive_Date == latest_date, Adult_Child == "Adult") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 10) %>%
  mutate(
    Specialty = fct_reorder(Specialty, Total),
    Pct_Long  = M18_Plus / Total * 100
  )

ggplot(top10, aes(x = Specialty, y = Total, fill = Pct_Long)) +
  geom_col() +
  geom_text(aes(label = comma(Total)), hjust = -0.1, size = 3.8, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_gradient(low = "#41B6E6", high = "#E8491D",
                      name = "% waiting\n18+ months") +
  labs(
    title    = paste("Top 10 Outpatient Waiting Lists —", format(latest_date, "%B %Y")),
    subtitle = "Adults only | Bar colour = proportion waiting 18 or more months",
    x        = NULL, y = "Total Patients Waiting",
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "right",
        axis.text.y      = element_text(size = 11))
```

## Waiting List Composition Over Time

The stacked area chart below tracks how the total national waiting list is composed across the four time bands each month. The growing proportion of the **18+ months band** (red) relative to earlier bands is a key signal of systemic pressure accumulating over time — patients are not being cleared fast enough to prevent migration into the longest-wait categories.

```{r eda-stacked, fig.height=7}
band_monthly <- op_clean %>%
  filter(Adult_Child == "Adult") %>%
  group_by(Month) %>%
  summarise(
    across(c(M0_6, M6_12, M12_18, M18_Plus), \(x) sum(x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  pivot_longer(c(M0_6, M6_12, M12_18, M18_Plus),
               names_to  = "Band",
               values_to = "Patients") %>%
  mutate(Band = factor(Band,
                       levels = c("M0_6", "M6_12", "M12_18", "M18_Plus"),
                       labels = c("0–6 Months", "6–12 Months",
                                  "12–18 Months", "18+ Months")))

ggplot(band_monthly, aes(x = Month, y = Patients, fill = Band)) +
  geom_area(alpha = 0.9, colour = "white", linewidth = 0.2) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = c("#41B6E6", "#005EB8", "#003087", "#E8491D")) +
  labs(
    title    = "Outpatient Waiting List Composition by Time Band",
    subtitle = "National total | Adults | All specialties combined",
    x        = NULL, y = "Patients Waiting", fill = "Wait Band",
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x   = element_text(angle = 45, hjust = 1),
        legend.position = "right")
```

---

# Section 4: Time Series Analysis {.tabset}

## Overview

Understanding whether the waiting list is growing due to **long-term structural pressure**, **seasonal variation** or **irregular shocks** has fundamentally different policy implications. A purely seasonal increase (e.g. post-summer backlog) resolves differently to a sustained structural trend. STL decomposition separates these components, while the rolling average smooths month-to-month noise to reveal the underlying direction.

## STL Decomposition

STL (Seasonal-Trend decomposition using Loess) decomposes the national monthly total into three additive components: **trend** (long-run direction), **seasonal** (recurring calendar patterns), and **remainder** (unexplained variation). It is robust to outliers — particularly important given potential data disruptions in health administrative systems.

```{r ts-stl, fig.height=9}
national_ts_df <- op_clean %>%
  filter(Adult_Child == "Adult") %>%
  group_by(Month) %>%
  summarise(Total_Waiting = sum(Total, na.rm = TRUE), .groups = "drop") %>%
  arrange(Month)

national_ts <- ts(
  national_ts_df$Total_Waiting,
  start     = c(year(min(national_ts_df$Month)),
                month(min(national_ts_df$Month))),
  frequency = 12
)

stl_decomp <- stl(national_ts, s.window = "periodic", robust = TRUE)

trend_component <- stl_decomp$time.series[, "trend"]
trend_slope     <- coef(lm(trend_component ~ seq_along(trend_component)))[2]

autoplot(stl_decomp) +
  labs(
    title    = "STL Decomposition — National Outpatient Waiting List",
    subtitle = paste0("Estimated trend slope: +", round(trend_slope, 0),
                      " patients/month on average | Adults | All specialties"),
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 13)
```

> **Interpretation:** The trend slope of approximately **`r round(trend_slope, 0)` additional patients per month** quantifies the systemic growth pressure beyond seasonal variation. A positive remainder cluster at any point indicates a period of unexplained excess — potentially a data quality event or a genuine service disruption.

## Rolling Average & Month-on-Month Change

```{r ts-rolling, fig.height=7}
national_ts_df <- national_ts_df %>%
  mutate(
    Roll3   = rollmean(Total_Waiting, k = 3, fill = NA, align = "right"),
    MoM_Chg = (Total_Waiting - lag(Total_Waiting)) / lag(Total_Waiting) * 100
  )

p_trend <- ggplot(national_ts_df, aes(x = Month)) +
  geom_col(aes(y = Total_Waiting), fill = "#005EB8", alpha = 0.45) +
  geom_line(aes(y = Roll3), colour = "#E8491D", linewidth = 1.4, na.rm = TRUE) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title    = "National Outpatient Waiting List — Monthly Total with 3-Month Rolling Average",
    subtitle = "Blue bars = monthly total | Red line = 3-month rolling mean",
    x        = NULL, y = "Patients Waiting",
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_mom <- ggplot(national_ts_df %>% filter(!is.na(MoM_Chg)),
                aes(x = Month, y = MoM_Chg, fill = MoM_Chg > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("TRUE" = "#E8491D", "FALSE" = "#003087")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title    = "Month-on-Month % Change",
    subtitle = "Red = increase | Blue = decrease",
    x        = NULL, y = "% Change"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_trend / p_mom + plot_layout(heights = c(2, 1))
```

---

# Section 5: ARIMA Forecasting {.tabset}

## Overview

The most operationally valuable output of any waiting list analysis is a reliable forward projection. The NHIU and HSE capacity planners need to know not just where the list stands today, but where it is likely to be in 6–12 months under current conditions. `auto.arima()` is used with exhaustive (non-stepwise) model selection to find the best-fitting ARIMA structure for this series.

A **back-test** is run by withholding the last 6 months of data, fitting the model on the remaining history, and measuring forecast accuracy against the known values — providing an empirical estimate of how reliable future projections are likely to be.

## 12-Month Forecast

```{r arima-forecast, fig.height=7}
arima_model <- auto.arima(
  national_ts,
  stepwise      = FALSE,
  approximation = FALSE,
  seasonal      = TRUE
)

fc <- forecast(arima_model, h = 12, level = c(80, 95))

fc_df <- data.frame(
  Month    = seq(max(national_ts_df$Month) %m+% months(1),
                 by = "month", length.out = 12),
  Forecast = as.numeric(fc$mean),
  Lo80     = as.numeric(fc$lower[, 1]),
  Hi80     = as.numeric(fc$upper[, 1]),
  Lo95     = as.numeric(fc$lower[, 2]),
  Hi95     = as.numeric(fc$upper[, 2])
)

ggplot() +
  geom_ribbon(data = fc_df,
              aes(x = Month, ymin = Lo95, ymax = Hi95),
              fill = "#41B6E6", alpha = 0.25) +
  geom_ribbon(data = fc_df,
              aes(x = Month, ymin = Lo80, ymax = Hi80),
              fill = "#41B6E6", alpha = 0.40) +
  geom_line(data = national_ts_df,
            aes(x = Month, y = Total_Waiting),
            colour = "#003087", linewidth = 1.1) +
  geom_line(data = fc_df,
            aes(x = Month, y = Forecast),
            colour = "#E8491D", linewidth = 1.1, linetype = "dashed") +
  geom_vline(xintercept = max(national_ts_df$Month),
             linetype = "dotted", colour = "grey50") +
  annotate("text",
           x     = max(national_ts_df$Month),
           y     = max(national_ts_df$Total_Waiting) * 1.01,
           label = "Forecast →",
           hjust = -0.1, colour = "#E8491D", size = 4) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title    = "12-Month ARIMA Forecast — National Outpatient Waiting List",
    subtitle = paste0(
      "Model: ARIMA(", arima_model$arma[1], ",",
      arima_model$arma[6], ",", arima_model$arma[2], ")",
      " | Shaded bands = 80% and 95% confidence intervals"
    ),
    x        = NULL, y = "Patients Waiting",
    caption  = "Source: NTPF Open Data | Forecast: auto.arima()"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Back-Test Accuracy

```{r arima-backtest}
n_holdout      <- 6
train_ts       <- head(national_ts, length(national_ts) - n_holdout)
test_ts        <- tail(national_ts, n_holdout)
backtest_model <- auto.arima(train_ts, stepwise = FALSE, approximation = FALSE)
backtest_fc    <- forecast(backtest_model, h = n_holdout)

acc <- accuracy(backtest_fc, test_ts)

acc %>%
  as.data.frame() %>%
  rownames_to_column("Set") %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kable(caption = "Back-Test Forecast Accuracy — 6-Month Holdout") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

> **Interpretation:** MAPE (Mean Absolute Percentage Error) on the test set gives the most interpretable accuracy measure — a MAPE of 5% means the model's monthly forecasts are typically within 5% of the actual value. Given the short series length (< 3 years), wider-than-ideal confidence intervals are expected.

---

# Section 6: Specialty Clustering {.tabset}

## Overview

Not all specialties behave the same way. Some have large, stable waiting lists with gradual growth; others are small but growing rapidly; others are characterised by a very high proportion of long-term waiters. Treating them all identically in policy terms misses this structural variation.

K-means clustering groups specialties by **five behavioural features** derived from their waiting list history:

| Feature | Description |
|---|---|
| Mean Total | Average waiting list volume over the period |
| SD Total | Volatility — how much the list fluctuates |
| % 18m+ | Average proportion waiting over 18 months |
| % 0–6m | Average proportion of recent (short) waiters |
| Growth Rate | % change from first to last observation |

All features are **standardised (z-scored)** before clustering so that high-volume specialties do not dominate by scale alone.

## Feature Engineering

```{r cluster-features}
specialty_features <- op_clean %>%
  filter(Adult_Child == "Adult") %>%
  arrange(Archive_Date) %>%
  group_by(Specialty) %>%
  summarise(
    Mean_Total  = mean(Total,    na.rm = TRUE),
    Max_Total   = max(Total,     na.rm = TRUE),
    SD_Total    = sd(Total,      na.rm = TRUE),
    Pct_18Plus  = mean(M18_Plus / Total, na.rm = TRUE) * 100,
    Pct_0_6     = mean(M0_6 / Total,     na.rm = TRUE) * 100,
    First_Total = first(Total),
    Last_Total  = last(Total),
    N_Months    = n_distinct(Month),
    .groups = "drop"
  ) %>%
  mutate(
    Growth_Rate = ifelse(
      First_Total > 0,
      (Last_Total - First_Total) / First_Total * 100,
      NA_real_
    )
  ) %>%
  filter(N_Months >= 6, !is.na(Growth_Rate))

feature_matrix <- specialty_features %>%
  select(Mean_Total, SD_Total, Pct_18Plus, Pct_0_6, Growth_Rate) %>%
  scale()

rownames(feature_matrix) <- specialty_features$Specialty

cat("Specialties available for clustering:", nrow(specialty_features))
```

## Optimal k — Silhouette Analysis

The silhouette method evaluates cluster quality across a range of k values (2–8). For each k, the average silhouette width measures how well each specialty fits its assigned cluster versus the next-closest cluster. Higher is better; the optimal k is the value that maximises this width.

```{r cluster-silhouette, fig.height=6}
set.seed(42)
sil_widths <- map_dbl(2:8, function(k) {
  km  <- kmeans(feature_matrix, centers = k, nstart = 25)
  sil <- silhouette(km$cluster, dist(feature_matrix))
  mean(sil[, 3])
})

optimal_k <- which.max(sil_widths) + 1

ggplot(
  data.frame(k = 2:8, Silhouette = sil_widths),
  aes(x = k, y = Silhouette)
) +
  geom_line(colour = "#003087", linewidth = 1.1) +
  geom_point(size = 4, colour = "#003087") +
  geom_point(data = ~ filter(.x, k == optimal_k),
             colour = "#E8491D", size = 6) +
  annotate("text", x = optimal_k, y = sil_widths[optimal_k - 1] + 0.005,
           label = paste("Optimal k =", optimal_k),
           colour = "#E8491D", fontface = "bold", size = 4.5, vjust = -0.8) +
  labs(
    title    = "Silhouette Analysis — Optimal Number of Specialty Clusters",
    subtitle = "Higher silhouette width = better cluster separation",
    x        = "Number of Clusters (k)", y = "Mean Silhouette Width"
  ) +
  theme_minimal(base_size = 13)
```

## Cluster Visualisation

```{r cluster-viz, fig.height=8}
set.seed(42)
km_final <- kmeans(feature_matrix, centers = optimal_k, nstart = 50, iter.max = 100)

specialty_features <- specialty_features %>%
  mutate(Cluster = as.factor(km_final$cluster))

fviz_cluster(
  km_final,
  data         = feature_matrix,
  geom         = "text",
  ellipse.type = "convex",
  repel        = TRUE,
  palette      = c("#003087", "#005EB8", "#41B6E6", "#E8491D", "#FF8C00"),
  ggtheme      = theme_minimal(base_size = 12)
) +
  labs(
    title    = "Specialty Clusters — Waiting List Behaviour (PCA Projection)",
    subtitle = paste(optimal_k, "clusters | Axes = first two principal components of 5 features"),
    caption  = "Source: NTPF Open Data"
  )
```

## Cluster Profiles

```{r cluster-profiles}
cluster_profiles <- specialty_features %>%
  group_by(Cluster) %>%
  summarise(
    `N Specialties`   = n(),
    `Avg Total`       = round(mean(Mean_Total)),
    `Avg % 18m+`      = round(mean(Pct_18Plus), 1),
    `Avg Growth (%)`  = round(mean(Growth_Rate), 1),
    Specialties       = paste(sort(Specialty), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg % 18m+`))

cluster_profiles %>%
  select(-Specialties) %>%
  kable(caption = "Cluster Summary — Ranked by 18m+ Rate") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat("\nSpecialties by Cluster:\n")
cluster_profiles %>%
  select(Cluster, Specialties) %>%
  kable(caption = "Specialty Membership by Cluster") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

---

# Section 7: Anomaly Detection {.tabset}

## Overview

Anomaly detection identifies months where a specialty's waiting list deviated unexpectedly from its historical pattern. In health data science, anomalies serve a dual purpose: they can signal **genuine data quality issues** (e.g. a hospital failing to submit complete returns) or **real service disruptions** (e.g. the 2021 HSE cyberattack, consultant industrial action, or COVID-related backlogs). Distinguishing between these requires domain expertise — but automated flagging is the essential first step.

A **Z-score approach** is used: for each specialty, months where the waiting list falls more than 2.5 standard deviations from that specialty's own mean are flagged as anomalous.

## Anomaly Table

```{r anomaly-table}
anomaly_threshold <- 2.5

anomalies <- op_clean %>%
  filter(Adult_Child == "Adult") %>%
  group_by(Specialty) %>%
  mutate(
    Mean_Total = mean(Total, na.rm = TRUE),
    SD_Total   = sd(Total,   na.rm = TRUE),
    Z_Score    = (Total - Mean_Total) / SD_Total
  ) %>%
  ungroup() %>%
  filter(abs(Z_Score) > anomaly_threshold) %>%
  arrange(desc(abs(Z_Score))) %>%
  select(Archive_Date, Specialty, Total, Z_Score) %>%
  mutate(
    Z_Score      = round(Z_Score, 2),
    Direction    = ifelse(Z_Score > 0, "⬆ Spike", "⬇ Drop")
  )

if (nrow(anomalies) == 0) {
  cat("No anomalies detected at |Z| >", anomaly_threshold)
} else {
  anomalies %>%
    kable(caption = paste("Anomalies Detected (|Z| >", anomaly_threshold, ")")) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
    row_spec(which(anomalies$Z_Score > 0), background = "#fff3cd") %>%
    row_spec(which(anomalies$Z_Score < 0), background = "#f8d7da")
}
```

## Anomaly Plot — Top 5 Specialties

```{r anomaly-plot, fig.height=11}
top5_specs <- as.character(top10$Specialty[1:5])

anomaly_plot_data <- op_clean %>%
  filter(Adult_Child == "Adult", Specialty %in% top5_specs) %>%
  group_by(Specialty) %>%
  mutate(
    Mean_Total = mean(Total, na.rm = TRUE),
    SD_Total   = sd(Total,   na.rm = TRUE),
    Z_Score    = (Total - Mean_Total) / SD_Total,
    Is_Anomaly = abs(Z_Score) > anomaly_threshold
  ) %>%
  ungroup()

ggplot(anomaly_plot_data, aes(x = Archive_Date, y = Total)) +
  geom_line(colour = "#005EB8", linewidth = 0.9) +
  geom_point(data = filter(anomaly_plot_data, Is_Anomaly),
             aes(colour = "Anomaly"), size = 4) +
  scale_colour_manual(values = c("Anomaly" = "#E8491D"), name = NULL) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  facet_wrap(~ Specialty, scales = "free_y", ncol = 1) +
  labs(
    title    = paste("Anomaly Detection — Top 5 Specialties (|Z| >", anomaly_threshold, ")"),
    subtitle = "Red dots = months deviating more than 2.5 standard deviations from specialty mean",
    x        = NULL, y = "Patients Waiting",
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    strip.text      = element_text(face = "bold", size = 11)
  )
```

---

# Section 8: Chronic Disease Pressure Index {.tabset}

## Overview

The NHIU role specifically centres on the **Chronic Disease Management (CDM) dataset**. The five CDM-relevant specialties — Cardiology, Respiratory Medicine, Endocrinology, Neurology and Gastroenterology — are the primary areas of focus for Ireland's Sláintecare chronic disease programme.

To enable direct comparison and prioritisation across these five, a **Chronic Disease Pressure Index (CDPI)** is constructed as a weighted composite of three dimensions:

| Component | Weight | Rationale |
|---|---|---|
| Current volume | 40% | Scale of the problem — more patients = greater absolute burden |
| % waiting 18m+ | 35% | Severity — reflects how deeply embedded the backlog is |
| Month-on-month growth | 25% | Trajectory — a growing list today creates a larger problem tomorrow |

Each component is min-max normalised to 0–100 before weighting, so no single dimension dominates by scale.

## CDPI Results

```{r cdpi-table}
cdm_specialties <- c(
  "Cardiology", "Respiratory Medicine", "Endocrinology",
  "Neurology", "Gastro-Enterology"
)

cdpi <- op_clean %>%
  filter(Adult_Child == "Adult", Specialty %in% cdm_specialties) %>%
  group_by(Specialty) %>%
  arrange(Archive_Date) %>%
  summarise(
    Latest_Total = last(Total),
    Pct_18Plus   = last(M18_Plus) / last(Total) * 100,
    MoM_Growth   = (last(Total) - nth(Total, -2)) / nth(Total, -2) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    Vol_Score    = (Latest_Total - min(Latest_Total)) / (max(Latest_Total) - min(Latest_Total)) * 100,
    Long_Score   = (Pct_18Plus  - min(Pct_18Plus))   / (max(Pct_18Plus)  - min(Pct_18Plus))   * 100,
    Growth_Score = (MoM_Growth  - min(MoM_Growth))   / (max(MoM_Growth)  - min(MoM_Growth))   * 100,
    CDPI         = Vol_Score * 0.40 + Long_Score * 0.35 + Growth_Score * 0.25
  ) %>%
  arrange(desc(CDPI))

cdpi %>%
  select(Specialty, Latest_Total, Pct_18Plus, MoM_Growth, CDPI) %>%
  mutate(
    Latest_Total = comma(Latest_Total),
    Pct_18Plus   = round(Pct_18Plus, 1),
    MoM_Growth   = round(MoM_Growth, 2),
    CDPI         = round(CDPI, 1)
  ) %>%
  kable(
    caption  = "Chronic Disease Pressure Index — CDM Specialties",
    col.names = c("Specialty", "Current Total", "% 18m+", "MoM Growth (%)", "CDPI Score")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#E8491D", color = "white")
```

## CDPI Chart

```{r cdpi-chart, fig.height=6}
cdpi %>%
  pivot_longer(c(Vol_Score, Long_Score, Growth_Score),
               names_to = "Component", values_to = "Score") %>%
  mutate(
    Component = recode(Component,
                       Vol_Score    = "Volume (40%)",
                       Long_Score   = "18m+ Rate (35%)",
                       Growth_Score = "MoM Growth (25%)"),
    Specialty = fct_reorder(Specialty, CDPI)
  ) %>%
  ggplot(aes(x = Specialty, y = Score, fill = Component)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = c("#003087", "#005EB8", "#41B6E6")) +
  scale_y_continuous(limits = c(0, 310)) +
  geom_text(data = cdpi,
            aes(x = Specialty, y = 308, label = round(CDPI, 1), fill = NULL),
            hjust = 0, size = 4, fontface = "bold") +
  labs(
    title    = "Chronic Disease Pressure Index (CDPI)",
    subtitle = "Weighted composite: volume (40%) + 18m+ rate (35%) + MoM growth (25%)",
    x        = NULL, y = "Component Score", fill = "Component",
    caption  = "Source: NTPF Open Data | CDM specialties only"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

---

# Section 9: Health Equity — Adult vs Child {.tabset}

## Overview

Health equity analysis examines whether different population groups experience **systematically different access to care**. Paediatric patients represent a particularly important group — children experiencing long waits face developmental and educational consequences beyond the direct health impact. The NTPF has published separate adult and child reporting since April 2021, enabling this comparison.

Two dimensions are compared: **raw volume** (to understand scale) and **proportion waiting 18+ months** (to understand relative severity of access barriers).

## Equity Chart

```{r equity-chart, fig.height=9}
equity <- op_clean %>%
  filter(!is.na(Adult_Child)) %>%
  group_by(Month, Adult_Child) %>%
  summarise(
    Total      = sum(Total,    na.rm = TRUE),
    Long_Wait  = sum(M18_Plus, na.rm = TRUE),
    Pct_18Plus = Long_Wait / Total * 100,
    .groups = "drop"
  )

p_eq_total <- ggplot(equity, aes(x = Month, y = Total, colour = Adult_Child)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_colour_manual(values = c("Adult" = "#003087", "Child" = "#E8491D")) +
  labs(title = "Total Waiting List by Patient Group",
       x = NULL, y = "Patients Waiting", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_eq_pct <- ggplot(equity, aes(x = Month, y = Pct_18Plus, colour = Adult_Child)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = \(x) paste0(round(x, 1), "%")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_colour_manual(values = c("Adult" = "#003087", "Child" = "#E8491D")) +
  labs(title = "% of Waiting List — Waiting 18+ Months",
       x = NULL, y = "% of list in 18m+ band", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_eq_total / p_eq_pct +
  plot_annotation(
    title    = "Health Equity Analysis — Adult vs Paediatric Outpatient Waiting Lists",
    subtitle = "National totals | Source: NTPF Open Data",
    theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                     plot.subtitle = element_text(size = 11, colour = "grey40"))
  )
```

---

# Section 10: Statistical Significance Testing {.tabset}

## Overview

Observed differences in waiting list volumes between years could arise from genuine change or simply from natural variation. Statistical hypothesis testing distinguishes between the two.

A **Wilcoxon signed-rank test** is used rather than a paired t-test because:

1. Waiting list counts are right-skewed — normality cannot be assumed
2. The test is robust to outliers, which are common in health administrative data
3. The paired design controls for specialty-level differences (comparing the same specialty in 2024 and 2025)

**Cohen's d** is reported alongside the p-value because statistical significance alone does not convey the practical magnitude of the change — a technically significant result might still represent a clinically trivial difference.

## Test Results

```{r stat-test}
yoy_test <- op_clean %>%
  filter(Adult_Child == "Adult", Year %in% c(2024, 2025)) %>%
  group_by(Specialty, Month_Num, Year) %>%
  summarise(Total = sum(Total, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Year, values_from = Total, names_prefix = "Y") %>%
  filter(!is.na(Y2024), !is.na(Y2025))

wtest <- wilcox.test(
  yoy_test$Y2025, yoy_test$Y2024,
  paired      = TRUE,
  alternative = "greater",
  conf.int    = TRUE
)

cohens_d <- (mean(yoy_test$Y2025, na.rm = TRUE) - mean(yoy_test$Y2024, na.rm = TRUE)) /
             sd(c(yoy_test$Y2025, yoy_test$Y2024), na.rm = TRUE)

effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE                ~ "Large"
)

results_df <- data.frame(
  Metric = c("Test", "Pairs (n)", "W statistic", "p-value",
             "95% CI lower", "95% CI upper", "Cohen's d", "Effect size"),
  Value  = c(
    "Wilcoxon Signed-Rank (paired, one-sided)",
    nrow(yoy_test),
    round(wtest$statistic, 0),
    format.pval(wtest$p.value, digits = 4),
    round(wtest$conf.int[1], 0),
    round(wtest$conf.int[2], 0),
    round(cohens_d, 3),
    effect_label
  )
)

results_df %>%
  kable(caption = "Statistical Test: Waiting Lists 2025 vs 2024",
        col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(4, bold = TRUE, background = ifelse(wtest$p.value < 0.05, "#d4edda", "#f8d7da"))
```

```{r stat-interpretation}
if (wtest$p.value < 0.05) {
  cat("✅ Result: Waiting lists in 2025 were SIGNIFICANTLY HIGHER than 2024 (p <",
      round(wtest$p.value, 4), ")\n")
  cat("   Effect size:", effect_label, "(Cohen's d =", round(cohens_d, 3), ")")
} else {
  cat("ℹ️  Result: No statistically significant increase from 2024 to 2025 (p =",
      round(wtest$p.value, 4), ")")
}
```

---

# Section 11: Regression — Growth Drivers {.tabset}

## Overview

While time series analysis describes *how* the national total moves, regression analysis asks *what drives* the total at the specialty level. By modelling `Total` as a function of the four waiting time bands and a time index, we can quantify the **relative contribution of each band** to overall waiting list size.

This matters for intervention design: if the 18+ months band is the dominant driver, interventions should focus on clearing the chronic long-wait backlog. If the 0–6 months band dominates, the issue is high incoming demand that must be managed through access reform.

## Regression Results

```{r regression, fig.height=6}
reg_data <- op_clean %>%
  filter(Adult_Child == "Adult") %>%
  mutate(Time_Index = as.numeric(difftime(Archive_Date,
                                          min(Archive_Date),
                                          units = "days")))

lm_model <- lm(
  Total ~ M0_6 + M6_12 + M12_18 + M18_Plus + Time_Index,
  data = reg_data
)

tidy_lm <- tidy(lm_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Significant = p.value < 0.05,
    term = recode(term,
                  M0_6       = "0–6 Months",
                  M6_12      = "6–12 Months",
                  M12_18     = "12–18 Months",
                  M18_Plus   = "18+ Months",
                  Time_Index = "Time Trend")
  )

tidy_lm %>%
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high, Significant) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(
    caption   = paste("OLS Regression Results | R² =", round(glance(lm_model)$r.squared, 4)),
    col.names = c("Term", "Estimate", "Std Error", "t", "p-value",
                  "CI Low", "CI High", "Significant")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(tidy_lm$Significant), bold = TRUE, background = "#d4edda")
```

## Coefficient Plot

```{r regression-plot, fig.height=6}
ggplot(tidy_lm, aes(x = estimate, y = reorder(term, estimate),
                    colour = Significant)) +
  geom_point(size = 5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.25, linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(
    values = c("TRUE" = "#003087", "FALSE" = "#E8491D"),
    labels = c("TRUE" = "Significant (p < 0.05)", "FALSE" = "Not significant"),
    name   = NULL
  ) +
  labs(
    title    = "Regression Coefficients — Drivers of Total Waiting List Size",
    subtitle = paste0("OLS | R² = ", round(glance(lm_model)$r.squared, 3),
                      " | Error bars = 95% confidence intervals"),
    x        = "Coefficient Estimate", y = NULL,
    caption  = "Source: NTPF Open Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

# Section 12: Policy Summary {.tabset}

## Overview

The final output is a clean, stakeholder-ready summary table of the kind that would be shared with the **HSE CDM programme team**, the **Department of Health**, or the **Sláintecare Implementation Office**. Each specialty is assigned a priority flag based on the proportion of its waiting list that has been waiting 18 months or more — the metric most closely associated with health equity and patient harm from delayed diagnosis.

| Flag | Threshold | Interpretation |
|---|---|---|
| 🔴 Critical | > 40% waiting 18m+ | Immediate systemic intervention required |
| 🟠 High | 25–40% waiting 18m+ | Targeted commissioning and validation review needed |
| 🟡 Moderate | 10–25% waiting 18m+ | Monitor closely; proactive outreach to patients |
| 🟢 Low | < 10% waiting 18m+ | Within manageable range; standard oversight |

## Priority Table

```{r policy-table}
policy_summary <- op_clean %>%
  filter(Adult_Child == "Adult", Archive_Date == latest_date) %>%
  arrange(desc(Total)) %>%
  mutate(
    Pct_18Plus    = round(M18_Plus / Total * 100, 1),
    Pct_0_6       = round(M0_6     / Total * 100, 1),
    Priority_Flag = case_when(
      Pct_18Plus > 40 ~ "🔴 Critical",
      Pct_18Plus > 25 ~ "🟠 High",
      Pct_18Plus > 10 ~ "🟡 Moderate",
      TRUE            ~ "🟢 Low"
    )
  ) %>%
  select(Specialty, Total, M0_6, M6_12, M12_18, M18_Plus,
         Pct_18Plus, Pct_0_6, Priority_Flag) %>%
  rename(
    `0–6m`      = M0_6,
    `6–12m`     = M6_12,
    `12–18m`    = M12_18,
    `18m+`      = M18_Plus,
    `% 18m+`    = Pct_18Plus,
    `% 0–6m`    = Pct_0_6,
    `Priority`  = Priority_Flag
  )

policy_summary %>%
  kable(
    caption = paste("Outpatient Waiting List Priority Table —",
                    format(latest_date, "%B %Y")),
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 12
  ) %>%
  row_spec(which(policy_summary$Priority == "🔴 Critical"),
           bold = TRUE, background = "#f8d7da") %>%
  row_spec(which(policy_summary$Priority == "🟠 High"),
           background = "#fff3cd")
```

---

# Summary & Conclusions

This analysis of NTPF Outpatient Waiting List data (2024–2026) demonstrates an end-to-end health data science workflow applied to a dataset of direct relevance to the HSE National Health Intelligence Unit.

**Key findings:**

- The national outpatient waiting list has grown at approximately **`r round(trend_slope, 0)` additional patients per month** on average, representing a structural rather than seasonal pressure
- **Cardiology** carries the highest absolute burden and one of the highest proportions of long waiters, making it the most critical specialty for CDM-aligned intervention
- The **18+ months band** has grown as a proportion of the total over the study period, indicating that newly arriving patients are not displacing the existing long-wait backlog
- Year-on-year statistical testing confirms the growth from 2024 to 2025 was **statistically significant** (Wilcoxon signed-rank, p `r ifelse(wtest$p.value < 0.001, "< 0.001", paste("=", round(wtest$p.value, 4)))`)
- K-means clustering reveals that specialties fall into distinct behavioural groups — a finding with direct implications for how commissioning and validation resources should be targeted

**Methodological notes:**

All code is reproducible and pulls directly from live NTPF CSV endpoints. The analytical pipeline — ETL, quality assessment, EDA, time series modelling, clustering, anomaly detection, significance testing and policy output — reflects the kind of end-to-end workflow expected of a Grade VII Health Data Scientist working across the NHIU's health data science programmes.

---

*Report generated `r format(Sys.time(), "%d %B %Y at %H:%M")` | Data source: [NTPF Open Data](https://www.ntpf.ie/waiting-list-data/open-data/) | Analysis: R `r R.version$major`.`r R.version$minor`*
